Supplementary Material for ``Provisional Title''

### 2021-04-15 draft/supp.tex
aesthetic changes (for consistency and clarity of math notations)

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{mathtools} % nice matrices
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
%% \usepackage[hidelinks]{hyperref} % hyperlinks
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xcolor} % text color
\usepackage[margin=3cm]{geometry} % margins

@@

Performing 16S rRNA gene sequencing on such a community yields a list of
sequences (exact sequence variants or ESVs) and their respective abundance.
Normalizing by the total number of sequences, we can obtain a vector $\mathbf{s}
= \left( s_1, s_2, \dotsc \right)$ that also satisfies the normalization
conditions of equations \ref{supp-eqn:normx_1} and \ref{supp-eqn:normx_2}.
Only if there are as many ESVs as species $\mathbf{s}$ has length $N$
(the same as as $\mathbf{x}$), but in the most general case it is possible that
a) multiple species share a same ESV and/or b) species carrying multiple copies
of the 16S rRNA gene have different sequences for each copy\mr. We denote
the length of $\mathbf{s}$ as $M$, which need not be equal to $N$.

@@

Having built our $\mathbf{Q}$ matrix according to equation
\ref{supp-eqn:Q_modified}, obtaining species abundances from sequencing data
is relatively straigthforward: we need to find the $\mathbf{x}$ that satisfies
equation \ref{supp-eqn:Qxs} for a given $\mathbf{s}$. However, in some cases
sequencing noise can introduce deviations in $\mathbf{s}$ that make it so the
vector $\mathbf{x}$ that solves equation \ref{supp-eqn:Qxs} does not meet the
boundary conditions in equation \ref{supp-eqn:normx_2}, i.e., some $x_i$ may be
negative or greater than 1. These cases are obviously problematic, but we can
avoid them by accounting for potential deviations in $\mathbf{s}$. We do this
by solving the following nonlinear optimization problem. We want to obtain an estimate of the true community composition $\mathbf{x}$
(that we will denote as $\mathbf{\hat{x}}$) so that the product
$\mathbf{Q} \cdot \mathbf{\hat{x}}$ is as close as possible to $\mathbf{s}$ while
having $\mathbf{\hat{x}}$ satisfy the normalization condition in equation
\ref{supp-eqn:normx_1} and the boundary conditions in equation
\ref{supp-eqn:normx_2}. For every possible $\mathbf{\hat{x}}$, we define a vector
$\mathbf{\epsilon} = \mathbf{\epsilon} \left( \mathbf{\hat{x}} \right) = 
\left( \epsilon_1, \epsilon_2, \cdots, \epsilon_M \right)$ as

\begin{equation}
\mathbf{\epsilon} \left( \mathbf{\hat{x}} \right) =
\mathbf{Q} \cdot \mathbf{\hat{x}} - \mathbf{s}
\end{equation}
%
and a function $f \left( \mathbf{\hat{x}} \right)$ as

\begin{equation}
f \left( \mathbf{\hat{x}} \right) = \sum_i \epsilon_i^2
\end{equation}
%
We also define a function $h \left( \mathbf{\hat{x}} \right)$ as

\begin{equation}
h \left( \mathbf{\hat{x}} \right) = 1 - \mathbf{\hat{x}}
\end{equation}
%
The best estimation for the species composition of a community will be the
$\mathbf{\hat{x}}$ that minimizes $f \left( \mathbf{\hat{x}} \right)$ while
satisfying the normalization condition (eq. \ref{supp-eqn:normx_1}) and the
boundary conditions (eq. \ref{supp-eqn:normx_2}). We solve
this problem using the \texttt{nloptr} package for R\mr, passing it
$f \left( \mathbf{\hat{x}} \right)$ as the function to minimze and
$h \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint, as well as
0 and 1
as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$. Further
details can be found in the code for the analysis (see section 
\nameref{supp-datacode} of this Supplementary Material).

@@

\subsection{Linear mixed-effects model trees}
\label{suppl-linear-mixed-model-results}

@@

% references
{\small
  %% \bibliography{refs}
  \bibliography{extracted-supp}
% \bibliographystyle{unsrt}
%% \bibliographystyle{mystyle}
\bibliographystyle{mybibwithurl}
}

>>>>

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{mathtools} % nice matrices
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xcolor} % text color
\usepackage[margin=3cm]{geometry} % margins

@@

Performing 16S rRNA gene sequencing on such a community yields a list of
sequences (exact sequence variants or ESVs) and their respective abundance.
Normalizing by the total number of sequences, we can obtain a vector $\mathbf{s}
= \left( s_1, s_2, \dotsc \right)$ that also satisfies the normalization
conditions of equations \ref{supp-eqn:normx_1} and \ref{supp-eqn:normx_2}.
Only if there are as many ESVs as species does $\mathbf{s}$ have length $N$
(the same as $\mathbf{x}$), but in the most general case it is possible that
\textit{a)} multiple species share a same ESV and/or \textit{b)} species carrying
multiple copies
of the 16S rRNA gene have different sequences for each copy\mr. We denote
the length of $\mathbf{s}$ as $M$, which need not be equal to $N$.

@@

Having built our $\mathbf{Q}$ matrix according to equation
\ref{supp-eqn:Q_modified}, obtaining species abundances from sequencing data
is relatively straigthforward: we need to find the $\mathbf{x}$ that satisfies
equation \ref{supp-eqn:Qxs} for a given $\mathbf{s}$. However, in some cases
sequencing error can introduce deviations in $\mathbf{s}$ that make it so the
vector $\mathbf{x}$ that solves equation \ref{supp-eqn:Qxs} does not meet the
boundary conditions in equation \ref{supp-eqn:normx_2}, i.e., some $x_i$ may be
negative or greater than 1. These cases are obviously problematic, but we can
avoid them by accounting for potential deviations in $\mathbf{s}$. We do this
by solving the following nonlinear optimization problem: we want to obtain an
estimate of the true community composition
(we will denote the estimate as $\mathbf{\hat{x}}$) so that the product
$\mathbf{Q} \cdot \mathbf{\hat{x}}$ is as close as possible to the $\mathbf{s}$
obtained from sequencing while
having $\mathbf{\hat{x}}$ satisfy the normalization condition in equation
\ref{supp-eqn:normx_1} and the boundary conditions in equation
\ref{supp-eqn:normx_2}. For every possible $\mathbf{\hat{x}}$, we can define a vector
$\boldsymbol{\upvarepsilon} = \boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) = 
\left( \varepsilon_1, \varepsilon_2, \cdots, \varepsilon_M \right)$ as

\begin{equation}
\boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) =
\mathbf{Q} \cdot \mathbf{\hat{x}} - \mathbf{s}
\end{equation}
%
and a function $f \left( \mathbf{\hat{x}} \right)$ as

\begin{equation}
f \left( \mathbf{\hat{x}} \right) =
\left| \boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) \right| ^2 =
\sum_i \varepsilon_i^2 \left( \mathbf{\hat{x}} \right)
\end{equation}
%
We can also define a function $\mathbf{h} \left( \mathbf{\hat{x}} \right)$ as

\begin{equation}
\mathbf{h} \left( \mathbf{\hat{x}} \right) = 1 - \mathbf{\hat{x}}
\end{equation}
%
The best estimation for the species composition of a community will be the
$\mathbf{\hat{x}}$ that minimizes $f \left( \mathbf{\hat{x}} \right)$ while
satisfying the normalization condition (eq. \ref{supp-eqn:normx_1}) and the
boundary conditions (eq. \ref{supp-eqn:normx_2}). We solve
this problem using the \texttt{nloptr} package for R\mr, passing it
$f \left( \mathbf{\hat{x}} \right)$ as the function to minimize,
$\mathbf{h} \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint,
and 0 and 1
as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$. Further
details can be found in the code for the analysis (see section 
\nameref{supp-datacode} of this Supplementary Material).

@@

\subsection{Example subsection}
\label{supp-results:example}

@@

% references
%{\small
  %% \bibliography{refs}
%  \bibliography{extracted-supp}
% \bibliographystyle{unsrt}
%% \bibliographystyle{mystyle}
%\bibliographystyle{mybibwithurl}
%}
